---
title: "HW 6: Hypothesis Tests"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes some helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
# You will need to install it (once) from GitHub.
# library(devtools)
# devtools::install_github("physicsland/ezids")
# Then load the package in your R session.
library(ezids)
```


```{r setup, include=FALSE}
# Some of common RMD options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
knitr::opts_chunk$set(results="markup", warning = F, message = F)
# Can globally set option for number display format.
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
```

```{r}
# 1. Do not provide answers/comments inside code blocks (like here) -- those are notes between coders/self and will be ignored for grading. 
# 2. Make sure your knitr options are set to include all results/code to be graded in the final document.
# 3. All charts/graphs/tables should have appropriate titles/labels/captions. 
# 4. Compose your answers using inline R code instead of using the code-block output as much as you can. 
# 5. Your grade is also determined by the style. Even if you answer everything correctly, but the .html does not look appealing, you will not get full credit. Pay attention to the details that we mentioned in class/homework and in previous sample .Rmd files. For example, how to use #, ##, ###, ..., bold face, italics, inline codes, tables, ..., {results = "asis"}, use of colors in plots/ggplots, and so forth.
```

This exercise uses the same graduate school admissions dataset as HW5.  The dataset is called `gradAdmit` on api.regression.fit, or as a local file *LogRegAdmit.csv*.  

The variables in the dataset are:  

* `admit`: 0 or 1 (False or True)
* `gre`: gre score
* `gpa`: grade point average at undergraduate level
* `rank`: the ranking of the applicant's undergraduate institute, 1, 2, 3, or 4

# Question 1  
**Read in the dataset and make sure the data type for the variables are set properly as numeric or factor variables.**


# Question 2
**Using the `outlierKD2()` function, identify and flag as NA the outliers for `gre` and `gpa`. Save the resulting data frame as a new data frame. This will require multiple lines of code -- you need to remove outliers for one variable at a time. How many NA values are in `gre` and in `gpa`?**


# Question 3
**Separate the data frame created in Q2 into two subsets: one for admitted and one for rejected.  How many observations in each data frame? **


# Question 4
**Now check for normality on the numerical variables. Produce histograms (use `ggplot()`) and QQ-plots (use `qqnorm()`) to check the distributions of the quantitative variables for the admitted and rejected subsets. Make a brief comment on the results.**


# Question 5
**Do the two subsets have different `gre` averages and `gpa` averages? Use the standard $\alpha$ = 0.05 to carry out a hypothesis test for differences in means. Be sure to state the null and alternative hypotheses, p-values, and conclusions from the tests.**


# Question 6
**Using the dataset from Q2, construct a contingency table between `rank` and `admit`. Are these two variables qualitative or quantitative?**


# Question 7
**Carry out a test to determine if `rank` and `admit` are independent, according to the contingency table above.  Use the standard $\alpha$ = 0.05. Be sure to state the null and alternative hypothesis, p-value, and conclusions from the test.**


# Question 8
**Using the dataset from Q2, test whether students from the four ranks have the same average `gre` or not. And also test if they have the same average `gpa` or not. Be sure to state the null and alternative hypotheses, p-values, and conclusions from the tests.  Remember that if they are not all the same, you will need to follow up with a post hoc test. Make brief comments on your results.**







